---
title: "Big Data Analytics Week 0"
author: "Ulrik Lyngs (based on Bertie Vidgen's materials)"
date: "2020-02-17"
output:
  html_document: default
---
 
# BDA Week 0 notes

## Overview
This a technical primer that doesn't contain anything you will be assessed on, but is for us to touch on a few technical issues before we get into the content.
If you have any questions please just ask now or email me at ulrik.lyngs@cs.ox.ac.uk

We will run over:
- R, RStudio and R Packages
- R Markdown
- ggplot2
- the 'rBDA' package
- Some basic stats
- Google and Stack Overflow
- Knitting the code into an html document


### Setup
The document you are looking at is an 'R Markdown' document, as we will explain below.
Each week's materials has an R Markdown file, which contains the practical examples that we will walk through together in the lecture.

Every week's R Markdown file begins with a 'setup' code chunk like the below - it loads in the packages, data and other options for the entire script.

```{r setup}
# create a function to check if a list of packages are installed, and install them if they aren't
fun_check_packages <- function(x){
  for (i in seq(1,length(x))){
    if (!x[i] %in% installed.packages()){
    install.packages(x[i], dependencies = T)}
    library(x[i],character.only=TRUE)}
}

# check if the tools we need are installed
packs <- c('dplyr','ggplot2', 'knitr', 'ggthemes','devtools', 'gridExtra')
fun_check_packages(packs)

options(scipen = 10) # makes the output more readable by increasing the number of values before scientific notation is used.
```


### R, RStudio and R Packages
R is a free statistical coding language. If you're unfamiliar with it, a fantastic overview is available here - https://cran.r-project.org/doc/contrib/Paradis-rdebuts_en.pdf

RStudio is an Integrated Development Environment (IDE) for R - that is, it is an application that combines a text editor, a terminal for running code, a file browser, a viewing area for seeing our plots and images, and more. It is what we will use for this course as it is way more accessible than R on its own.

R packages are downloadable packages which contain extra functions beyond those available in the Base R installation. Most of the well-used R packages we use are available on 'CRAN', a network of servers around the world that store code and documentation for R. These can be downloaded and loaded easily:

```{r package-1}
#install.packages('dplyr') #install package
library(dplyr) #load package
```

Other packages that we use are not available in CRAN but can be collected from GitHub using a package called `devtools`. First, you need to install devtools - with the command `install.packages("devtools")` - and then you can install packages from GitHub repositories.

```{r package-2}
#install.packages("devtools")
devtools::install_github("bvidgen/rBDA")
library(rBDA) #you can use library with a github package just as you would with a CRAN pakage
```

You can then use the functions in this package, the same as you would with packages from CRAN.

One risk in loading many packages at once is that they will have functions with the same names. Obviously, this is not ideal as R won't know which function you are trying to use. The issue can be resolved by using the :: syntax, which makes it explicit which package a function comes from.

```{r package-3}
df <- data.frame('x' = rep(1:2,50))
df


nrow(dplyr::distinct(df))

# you can also use the :: to invoke base R functions:
nrow(base::unique(df))

```

If you are using a function and you can't remember what input params or optional arguments it takes you can check by using the ? syntax:

```{r package-4}
?mean
?dplyr::distinct
```

#### RStudio Projects
The materials for each week contain a 'Project' file, that is, a file ending in the extension .Rproj.

Using RStudio projects is best practice for making sure R (and yourself) knows where to look for files. Basically, when you open an RStudio project file, R's working directory is automatically set to wherever folder on your computer contains that project file. This makes it much easier to share your work with other people, because you won't be trying to read in files using a filepath that is effectively `~/path/that/only/exist/on/my/computer.csv`.

Every week, open RStudio by opening the R Project file in the materials for that week - then, you should be able to run the content in the R Markdown file without any trouble. 

### R Markdown
R Markdown is a fantastic tool that allow us to mix normal text written in the very simple 'markdown' syntax (see https://en.wikipedia.org/wiki/Markdown) with chunks of code. It's a bit like Jupyter notebooks, that people use for Python, but only better...

An R Markdown document is a plain text file that you can open in any text editor. When you open it in RStudio, you can run the chunks of code and use it as an interactive notebook. You can also export ('knit') the document to other formats, such as a Microsoft Word or HTML document, using the 'Knit' button. (You could write your homework assignments this way!)

To run code one line at a time, click on a line of code and press 'cmd' + 'return'. To run all the code in a chunk, press 'cmd', 'shift' and 'return', or click the green 'play' button on the top right in a code chunk.

When you knit your code to, for example, an HTML file, you can control the output. If you want to have all of the input and the output in your knitted document then leave it as it is:

```{r knit-1}
x <- 123
print(x)
```

With warnings:

```{r knit-2}
print(log(-1))
```

Without warnings (this is quite useful with graphics):

```{r knit-3, warning = FALSE}
print(log(-1))
```

And to hide everything (but still have the objects created in the environment):

```{r knit-5, include = FALSE}
x <- 456
for (i in 1:10){
  print(i)}
print(x)
```


### ggplot2
ggplot2 is the main graphics package we will use during the course. It was created by Hadley Wickham who is responsible for what is called the 'tidyverse' - an awesome suite of R packages for data science https://www.tidyverse.org

ggplot2 is based on the 'Grammar of Graphics' and lets you make nice looking graphs easily (if you've taken Info Viz then you'll be familiar with this). Most features of graphs can be customized.

Here's a walkthrough for building up a ggplot2 graph - use this as resource for the assignments.

With ggplot2 you want to create layers of graphics, this is the 'grammar of graphics', as described here: https://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf - we will touch on this more in the future weeks.

```{r ggplot1, echo=T}
# generate some test data
df <- data.frame(first_variable = seq(1,100, by = 7),
                second_variable = seq(1,100, by = 7))

ggplot2::ggplot(df, aes(x = first_variable, y = second_variable))
# the first argument is the dataframe - here it is 'df' as that is the name of our dataframe
# the second and third arguments are in aes() which stands for 'aesthetic' - these are the variables you want to plot

# Running the line above makes an empty graph. We need to add in the geometry we want to plot
ggplot(df, aes(x = first_variable, y = second_variable)) +
  geom_line()

# we could also plot other geometries
ggplot(df, aes(x = first_variable, y = second_variable)) +
  geom_point()

# or several at once
ggplot(df, aes(x = first_variable, y = second_variable)) +
  geom_line() +
  geom_point()

# we can adjust many arguments for each type of geometry, such as color and size
ggplot(df, aes(x = first_variable, y = second_variable)) +
  geom_line(color = 'red', size = 10) + 
  geom_point(color = 'yellow', size = 1) # tho, obviously, never actually make a graph this ugly...

```

It is possible to plot two graphs at the same time (using data from a single dataframe) using (i) facet wrap and (ii) a third factor variable

```{r ggplot2, echo=T}
# generate data
df1 = data.frame(x = rep(seq(1,100),2), 
                   y = c(seq(1,100),seq(100,1)),
                 z = c(rep('a', 100), rep('b', 100)))

# plot using facet wrap
ggplot(df1, aes(x,y)) +
  facet_wrap(~z) +
  geom_line()
    # facet wrap splits the data into separate graphs based on a third variable in the dataframe (here, 'z')
    # it is incredibly useful for subsetting by a variable and producing many graphs at once

# Alternatively, we can invoke a third variable as a factor and plot the two lines on one graph
ggplot(df1, aes(x,y, color = z)) +
  geom_line()
```

There's lots of options to adjusting your plots in ggplot2

```{r ggplot3, echo=T}
# we can adjust the axis scales on ggplot graphs easily using xlim() and ylim()
ggplot(df, aes(first_variable, second_variable)) +
  geom_line(color = 'blue',
            size = 1) +
  xlim(0, 1000) + 
  ylim(0,250) #obviously, this actually looks quite stupid...

# and we can adjust the x and y axis labels and add a title
ggplot(df, aes(first_variable, second_variable)) +
  geom_line(color = 'blue',
            size = 1) +
  labs(x = "some variable", y = "another variable", title = "a plot to show how to use ggplot2")
      # As a warning - often ggplot titles look a bit odd when you paste the graphic elsewhere, such as into word. Sometimes it's better to not put a title onto the graph itself and just write one out in the text.

# annotate the graph
ggplot(df, aes(first_variable, second_variable)) +
  geom_line(color = 'blue',
            size = 1) +
  labs(title = "a plot to show how to use ggplot2") +
  annotate("text",
           x = 25, 
           y = 90, 
           label = "Something insightful", 
           size  = 8)

```

To make your graphics look nicer, and if you're super pernickety, you can manually adjust lots of parameters...

```{r ggplot4, echo=T}
ggplot(df, aes(first_variable, second_variable)) +
  geom_line(color = 'blue',
            size = 1) +
  labs(x = "some variable", y = "another variable", title = "a plot to show how to use ggplot2") +
  theme(title = element_text(size = 16,
                             face='bold'), # adjust the plot title
                 axis.text.y = element_blank(), # remove any values from the y axis
                 axis.text.x = element_text(color = 'black'), # change the x axis values text 
                 axis.title.x = element_text(size = 12,
                                             color = 'black',
                                             face = 'bold'), # change the x axis title
                 panel.grid.major = element_blank(), 
                 panel.grid.minor = element_blank(),
                 panel.border = element_blank(),
                 panel.background = element_blank()
                 )
```

And to save your output

```{r ggplot6, echo = T}
plot_out <- ggplot(df, aes(first_variable, second_variable)) +
  geom_line(color = 'blue',
            size = 1) +
  labs(title = "a plot to show how to use ggplot2") +
  annotate("text",
           x = 25, 
           y = 90, 
           label = "Something insightful", 
           size  = 8)

plot_out

ggsave("plot_out.pdf", 
       plot = plot_out,
       width = 20, 
       height = 10,
       units = 'cm') # adjust these values to produce a graph which is the right height / width
```

If there is something specific you want to do with ggplot2 then Google it - someone has most likely already tried to do it and posted about it on Stack Overflow.

Some people have also made extension packages for ggplot2. If you want to jazz up your plots you can use the ggthemes package (which we've already loaded). This lets you use some great (but also some horrific) looking themes in your graphs

```{r ggthemes}
df3 = data.frame(UK = rep(seq(1,100),2), 
                 US = c(seq(1,100),
                        rep(seq(1,20),5)),
                 happiness = c(rep('UK', 100),
                               rep('US',100)))
# FiveThirtyEight theme
ggplot(df3, 
       aes(UK, US, 
           color = happiness)) + 
  geom_line() + 
  ggthemes::theme_fivethirtyeight()

# Wall Street Journal theme
ggplot(df3, 
       aes(UK, US, 
           color = happiness)) + 
  geom_line() + 
  ggthemes::theme_wsj()

# Excel theme
ggplot(df3, 
       aes(UK, US, 
           color = happiness)) + 
  geom_line() + 
  ggthemes::theme_excel() # but don't actually use this as it looks like trash
```



### the rBDA package
rBDA is a package that Bertie Vidgen and Taha Yasseri specially created for the BDA class. There are four main things it enables you to do: 
(i) plot error bars on binned continuous data, 
(ii) calculate and plot autocorrelated and crosscorrelated values robustly, 
(iii) make panelled scatterplots and 
(iv) calculate values with a moving time window.

There are also some other functions which you can access but they aren't terribly exciting (like sd_pop which calculates population level standard deviation)

All we want to do today is just check that the functions are working. Run the commands below and after each one check the output. So long as you get *something* then it is probably working fine.

```{r BDA}
# generate test data
set.seed(1)
df <- data.frame(x = seq(1, 100),
                y = rnorm(100,50,10))

# generate some more test data
set.seed(101)
x <- rnorm(1000, mean=10, sd=10)
y <- x + rnorm(1000 ,mean=1, sd=5)
y <- c(y[901:1000], y[1:900]) 
df.test2 = data.frame(x, y); rm(x,y)


# 1. plot error bars
rBDA::error_bars(df)

# 2. CCF
# NOTE: this output will not 'knit' (see below), so comment out these three lines when knitting the document
out <- rBDA::cor_fun(df$x, df$y, lag.max = 10)
out$fit_best; out$wrap; out$conf.int
rBDA::cor_plot(out)

# 3. plot multiple scatter plots at once
rBDA::cor_scatter(df.test2$x,
                  df.test2$y,
                  list(50,100,250,500))

```

### Some basic stats
This course is pretty technical in places. Some things you want to have a good grasp of (all of which were covered in the term 1 stats course) are:
- covariance
- correlation
- logarithms
- linear regression (just simple, i.e. using one predictor)
- types of variables (i.e. ratio/interval)


### Google is your friend
Finally, just a comment on Google: 
learning how to Google effectively for advice and code is a real skill. Most problems you encounter will have been already tackled by someone else, and they will have posted it on Stack Overflow or publicly accessible email lists. Use these as much as you can to solve problems on your own. Though, that said, please do reach out if you ever have a problem as we are always on hand to help with issues.

A couple of tips for searching - be as precise as you can, use the technica/established terms where possible, and try to make your query broader than just your specific issue (i.e. don't write 'split a vector into 7 subsections of width 30' but write 'split a vector into n equally sized subsections')


### Knit the code
Finally, let's make an html document!

You can make an html document by using the 'Knit' command at the top of the RStudio interface, just below the script tabs. Knitting to html is quite straightforward. On the drop down next to 'Knit' set it to 'Knit to HTML'. The html is saved into the directory containing your .Rmd file. Notice that it will be saved to the folder where the R Project file lives, if you opened this R Markdown document by first opening the R Project file.

Knitting to Word is straightforward, though it does take a while. Just select 'Knit to Word' from the drop down menu.

Knitting to PDF is a bit more difficult. You need a TeX (as in LaTeX) installation on your computer. You might not want to bother with this unless you have some experience with LaTeX, but if you are up for it, you can: 
- try installing `TinyTex` which is made by the creator of R Markdown: https://yihui.org/tinytex/ 
- get a full installation of LaTeX from here: http://tug.org/mactex/
- use RStudio Cloud, which is running RStudio in a browser and might not require you to install anything at all: https://rstudio.cloud/

Alternatively, you could knit to word and export as a PDF from there.


*End of workshop notes*





